Resonant Scattering in a Uniform Sphere with Large Optical Depth

### 2021-09-24 ms.tex
Fixing formatting of appendix and changing instances of "line center optical depth" to \tau_0

Numerical simulations can be used to understand the complex layers of these irradiated exoplanet atmospheres. However, the extreme optical depths at \lya line center introduce a computational barrier to simulating radiative transfer via Monte Carlo methods. The main impediment to performing magnetohydrodynamic simulations with fully a coupled radiative transfer solution via Monte Carlo methods is the prohibitive computational cost. The number of scatterings a photon undergoes is proportional to the line center optical depth.  Near the base of the atmosphere, the line center optical depth is ${\sim}10^9$, so most of the time is spent following photons in these cells. A method that can treat these zones via a simple solution rather than Monte Carlo has the potential to greatly accelerate the calculation. A complete, general analytic solution for the resonance scattering has never been derived but approximate solutions exist in certain limits.

@@

Starting with the full transfer equation, Eq. \ref{app:rteqn_derivation}\ref{eq:finaleqn}, with no photon destruction and a photon emission term given by Eq. \ref{app:rteqn_derivation}\ref{eq:jem}, the steady-state transfer equation is
\be
\nabla^2 J + \left( \frac{k}{\Delta} \right)^2 \frac{\partial^2 J}{\partial \sigma^2} & = & 
- \frac{ \sqrt{6} kL}{4\pi \Delta^2} \delta^3(\vec{x} - \vec{x}_s) \delta (\sigma - \sigma_{\rm s}).
\label{eq:rt_no_destr}
\ee
where $\sigma$ comes from a change of variables in photon frequency (Eq. \ref{app:rteqn_derivation}\ref{eq:change_of_variables}) and $\sigma_s$ is the photon frequency of the source. The boundary condition is that there is no incoming intensity at the surface, which can be written
\citep{1986rpa..book.....R}
\be
J & = & \sqrt{3} H
\label{eq:bc}
\ee
at $r=R$.

@@

\subsection{Scaling with Line Center Optical Depth}
To examine the relative size of the correction term $H_{\rm bc}$ as compared to $H_0$, we consider the case where the $\tau_0$ is large and thus $J_{\rm bc} \gg H_{\rm bc}$. We obtain the proportionality
\be \label{eq:hbc_scaling}
\frac{H_{\rm bc}(R, \sigma)}{H_0(R, \sigma)} \propto \frac{1}{(a\tau_0)^{1/3}}.
\ee
Thus, the fractional importance of the correct frequency-dependent boundary condition term $H_{\rm bc}$ scales as $(a\tau_0)^{-1/3}$. At very large $\tau_0$, it is expected that the correction term will be small, but will become increasingly important as $\tau_0$ decreases.

The location of the spectral peak of the distribution is
\be \label{eq:tau_peak_scaling}
x_{\rm peak} {\sim} \left(a\tau_0\right)^{1/3}.
\ee
The solutions of the transfer equation will perform best when the peaks of the spectral energy distribution lie well outside of the Doppler core, e.g., for large $\tau_0$. For \lya and H atoms with $T=10^4$ K, the damping parameter is $a = \Gamma / (4\pi\Delta) = 4.72\times 10^{-4}$. Thus, the value of $x$ at which the Doppler and Lorentzian components of the line wing are equal is $x_{\rm cw}=3.3$. Setting $x_{\rm cw} = x_{\rm peak}$ and solving for $\tau_0$ gives the value of the line-center optical depth at which the core of the line moves into the peak of the spectrum, which is $\tau_{\rm cp}=7.3\times10^4$. Solutions derived assuming a Lorentzian line profile $\phi(x) \approx a/(\pi x^2)$ are expected to fail for lower line center optical depths where the peak of the spectrum lies near the core and improve for larger optical depths where the spectral peak is well outside of the line core.

\subsection{Comparison to Monte Carlo}
The Monte Carlo method is used to solve the transfer equation numerically. This method is valid at arbitrarily large optical depths, being restricted only by its computational demand which grows proportionally to the number of photons used and $\tau_0$. For each simulation, a total of ${\sim}10^6$ photon packets are initialized at a monochromatic source frequency $x_s$ and are allowed to propagate through the sphere until escaping, at which point their positions, outgoing angles, and escape frequencies are stored. A constant temperature of $T=10,000\ \rm K$ is set for the gas. Frequency redistribution is calculated at each scattering, including the effects of recoil, to obtain the spectrum at the surface of the spherical simulation domain. In the comparisons shown in this section, the raw photon data is binned in frequency to obtain spectra. 64 bins are used to ensure good statistics and minimal noise in each bin. Further details of the Monte Carlo implementation are discussed in \cite{2017ApJ...851..150H}.

\ifx
\begin{figure}
    \centering
    \includegraphics{taubc.pdf}
    \caption{Boundary condition term $H_{\rm bc}$ scaled by $(a\tau_0)^{1/3}$ shown for increasing line center optical depths. The Monte Carlo data minus the $H_0$ solution should be close to $H_{\rm bc}$ according to Eq. \ref{eq:totalflux}. As $\tau_0$ becomes larger, the agreement between these two improves, indicating that at large optical depths the relative size of the corrective boundary condition term scales as $(a\tau_0)^{-1/3}$}.
    \label{fig:taubc}
\end{figure}
\fi

@@

Figure \ref{fig:xinit} shows quantitative examination of how the boundary term $H_{\rm bc}$ scales with both $\tau_0$ and $x_s$, motivated by the previous figures' demonstration that $H_0$ appears to become more accurate as $\tau_0$ increases. In this figure, the photon initialization frequency $x_s$ is shifted by integer multiples of $\tau_0$ in $\sigma$ frequency units (Eq. \ref{app:rteqn_derivation}\ref{eq:change_of_variables}) such that the source, now written $\sigma_s$, falls near the peak of the spectrum for each $\tau_0$. For clarity, only the $x > 0$ side of the spectrum is shown. From Eq. \ref{eq:hbc_scaling}, it is expected that the size of $H_{\rm bc}$ should become smaller with larger optical depths following a $(a\tau_0)^{-1/3}$ scaling as compared to $H_0$. This factor has been scaled out of the figure such that solutions for different optical depths and the same source frequency $\sigma_s$ should show close agreement in scale on the figure's vertical axis if the relation holds. Indeed, the scaled solutions converge as the optical depth becomes larger, indicating general agreement with the $(a\tau_0)^{-1/3}$ scaling. The remaining discrepancy present in the vertical axis for fixed $\sigma_s$ is due to the width of the Doppler core with respect to the peak of the spectrum; at lower $\tau_0$, this causes the line profile approximation in the wing (Eq. \ref{app:rteqn_derivation}\ref{eq:line_profile_wing}) to break down. From this, we conclude that the errors introduced by the incorrect separation of variables in \cite{1973MNRAS.162...43H}, \cite{1990ApJ...350..216N}, \cite{2006ApJ...649...14D} and others are proportional to $(a\tau_0)^{-1/3}$.

@@

Let us define the damping frequency to be $\gamma \equiv i\omega$, which is real-valued and positive for damped solutions to the time-dependent equation. There are certain values of $\gamma$ for which a resonant response is produced which we will call $\gamma_{nm}$. Near these $\gamma_{nm}$ poles, the eigenfunctions will have the form 
\be \label{eq:jnsigmaomega_approx}
J(n,\sigma,-i\gamma) & \simeq \frac{ J_{nm}(\sigma) }{\gamma_{nm} - \gamma} + {\rm terms\ which\ vary\ slowly\ in\ \gamma}.
\ee
From Eq. \ref{eq:jrsigmat_expansion}, the contribution of a single pole to the total solution is
\be
J_{nm}(r,\sigma,t) & = &  -j_0(\kappa_n r)  \int \frac{d\gamma}{2\pi i} e^{-\gamma t}J(n,\sigma,-i\gamma)
\\ & \simeq & j_0(\kappa_n r)  \int \frac{d\gamma}{2\pi i} e^{-\gamma t} \frac{ J_{nm}(\sigma) }{\gamma - \gamma_{nm}}.
\ee
We close the integral along the imaginary $\gamma$-axis  with a semi-circle at infinity and apply the residue theorem. The contribution from the semi-circle is zero, so we obtain
\be
J_{nm}(r,\sigma,t) & \simeq & j_0(\kappa_n r) J_{nm}(\sigma) e^{-\gamma_{nm}t}.
\ee
Summing over all spatial modes $n$ and over all eigenmodes $m$ for a given $n$, we obtain
\be \label{eq:Jrsigmat}
J(r,\sigma,t) & = & \sum_{n=1}^\infty j_0(\kappa_n r)  \sum_{m=1}^{\infty} J_{nm}(\sigma) e^{-\gamma_{nm}t}.
\ee

% State that this works for some of the poles, but others are missing. Can't capture asymmetry in spectrum. There is some other term that's missing from this analysis.

%The divergent pieces don't contain the piece that can be asymmetric --- the poles we're finding don't allow asymmetry in the specturm.

% Missing something in our contour? Residue theorem doesn't apply in this case? Missing a pole or poles that are not present?

% We're losing some poles due to the boundary but we think that the sum of the decaying solutions is converging fast enough to cut it off somewhere.

% We try this ansatz for the symmetric case and it basically seems to work. We show this by numerical calculation. We have to point out that we've shown it doesn't work for an off-center source. But then what do we say? We don't really have an explanation for how it doens't work. WKB gives info about it, but we don't discuss thta in the paper either.

% After we do the symmetric case, note that due to what we found in the WKB, this does not capture the asymmetry in the source function. Leave it there and just. Don't say it's left for future work. Just leave it. Note that there's an issue and say "this suggests that our ansatz is incomplete in the asymmetric case."

% Change ordering of figures. Have current fig 8 and 9 be the last thing since they contain a sum over modes. Put solid results first.

@@

The form of a single eigenmode $J_{nm}(\sigma)$ is oscillatory out to some turning point, $\sigma_{tp}$, at which point the function becomes evanescent. The location of the turning point can be found by ignoring the delta-function discontinuity at the source frequency $\sigma_s$ in Eq. \ref{eq:diffusion_plugged_in} and re-examining the resulting homogeneous differential equation. We obtain
\be
\frac{d^2J}{d\sigma^2} & = & \left[ \left( \frac{\kappa_n \Delta }{k} \right)^2 - \frac{3\phi \gamma\Delta^2}{ck}\right] J,
\ee
where the line profile is approximated as in Eq. \ref{app:rteqn_derivation}\ref{eq:line_profile_wing}. When the coefficient on the right hand side is positive, exponential growth or decay solutions are found (evanescence). This occurs on the line wings. When the coefficient on the right hand side is negative, oscillatory solutions are found (propagation), which occurs near the line core. The boundary between propagation and evanescence occurs at the turning point, given by
\be
\sigma_{tp} & = & \sqrt{\frac{2a}{\pi}}\left( \frac{k \gamma}{ \kappa_n^2 c \Delta} \right)^{3/2}
\ee
Thus, to ensure accuracy in the final sum of Eq. \ref{eq:Jrsigmat}, the bounds of $\sigma$ must be set sufficiently far outside of $\sigma_{tp}$ such that the function is small at the edges. The scale of an $e$-folding in $J_{nm}(\sigma)$ is $k/(\kappa_n \Delta) = \tau_0 / (\sqrt{\pi} n)$, so a grid of $\sigma$ is chosen that spans a large enough number of $e$-foldings that no oscillatory behavior is present at the boundaries of the calculation.

@@

In Fig. \ref{fig:tau_scaling}, it is shown that the time constant of exponential decay in fitted Monte Carlo escape time distributions converges with $\gamma_{00}^{-1}$ at sufficiently high optical depths, following a $t\propto(a\tau_0)^{1/3}$ scaling. At lower optical depths $\tau_0$, the effects of line core scattering are most important, leading to a larger discrepancy in the characteristic escape timescale. Here, the Monte Carlo accurately characterizes the photons which scatter in the core many times before escaping, while the semi-analytic solution does not capture this behavior as it uses only the Lorentzian piece of the line profile and does not use enough spatial modes to accurately model the frequency regime near line center. However, as the optical depth grows, the effect of core scattering becomes smaller and the approximations hold, agreeing well with the expected $(a\tau_0)^{1/3}$ scaling for the rate constant.

\begin{figure}
    \centering
    \includegraphics[width=0.75\textwidth]{tau_scaling.pdf}
    \caption{Characteristic timescales as a function of optical depth. $\gamma_{00}^{-1}$ and a fit of the exponential falloff for the Monte Carlo escape time distribution are shown to converge to a $t \propto (a\tau_0)^{1/3}$ scaling at large optical depths. The Monte Carlo points on this figure are obtained by fitting only the exponential piece of the escape time distribution to obtain the rate constant.}
    \label{fig:tau_scaling}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics{waittime.pdf}
    \caption{A wait time distribution of escaping photons for two different line center optical depths $\tau_0$, which is fit using Eq. \ref{eq:fitting_function}. A sum over 20 spatial eigenmodes and 500 frequency eigenmodes is labeled ``Eigenfunctions''. All calculations were performed with a monochromatic source of photons at line center ($x_s = 0.0$).}
    \label{fig:escape_time}
\end{figure}

@@

Models of heating and escape of hydrogen gas in exoplanet atmospheres can be constructed with an accurate treatment of resonant scattering of \lya in spherical geometry. The Monte Carlo method can be used for this problem, but requires a solution of radiative transfer in a sphere and is limited by its high computational demand for large optical depths where there are many photon scatterings before escape. The primary goal of this work is to present a solution for mean intensity ($J$) and flux ($H = F/4\pi$) of photons near line-center frequency $\nu_0$ in a uniform sphere. We have generalized a solution derived by \cite{2006ApJ...649...14D} (called $H_0$ here) to allow a monochromatic source of photons with frequencies away from line center, and we have made a correction to their incorrect $J=0$ surface boundary condition. This correction allows the proper condition $J=\sqrt{3}H$ to be satisfied at the surface of the sphere, and comes from the introduction of the terms $J_{\rm bc}$ and $H_{\rm bc}$ which are solved using a continuous Fourier expansion in frequency. Enforcement of the correct boundary term leads naturally to expansion of the spatial dependence in terms of spherical Bessel functions. In practice, the continuous expansions are discretized and the Fourier coefficients solved for via a matrix equation. The resulting flux correction, $H_{\rm bc}$, scales as $(a\tau_0)^{-1/3}$. Thus, for very large $\tau_0$, only a small correction to $H_0$ is needed, while larger errors are present in calculations performed at lower $\tau_0$. Since the Laplacian form of the differential equation is only attainable by approximating the line profile as $\phi \approx a/(\pi x^2)$, these solutions are invalid in areas where the Doppler core of the \lya line has a significant effect on the scattering of photons. The peak of the spectral energy distribution of escaping photons is $x_0 = (a\tau_0)^{1/3}$; as such, calculations performed at low optical depths are likely to be inaccurate due to the close proximity of the spectral peak and the Doppler core of the line. By comparison with Monte Carlo simulations, we have shown that the enforcement of the correct frequency-dependent boundary conditions improves the accuracy of these analytic solutions %and correctly tracks changes in the spectrum of escaping photons as the monochromatic source frequency $x_s$ is altered, 
provided a sufficiently large $\tau_0$. Specifically, this solution shows improvement over previous solutions that utilized a $J=0$ surface boundary condition, such as \cite{1973MNRAS.162...43H}, \cite{1990ApJ...350..216N}, \cite{2006ApJ...649...14D} and others. TODO: Give details. The surface boundary condition correction $H_{\rm bc}$ is positive in the line wing and negative at the peak of the spectrum, which suggests the previously cited solutions are too large at the spectral peaks and too small further out in the wing, where the error scales approximately as $(a\tau_0)^{-1/3}$. This appears to be the consequence of not enforcing the correct frequency-dependent boundary conditions at the surface of the medium.

@@

We have examined previous solutions to \lya transfer in the limit of large optical depth, noting that the separation of variables and treatment of the boundary condition in \cite{1973MNRAS.162...43H}, \cite{1990ApJ...350..216N}, \cite{2006ApJ...649...14D} and others does not produce a valid solution to the transfer equation. Here, we have derived the solution in spherical geometry with an appropriate treatment of the surface boundary condition as a frequency-dependent quantity, examining the spectral and time dependence of the resulting equations. The key result is that the errors in the previously-cited works have been quantified via a corrective boundary term, $H_{\rm bc}$, which resolves an excess in flux at the spectral peak and a deficit in the line wing of the calculated spectrum of \lya radiation as compared with Monte Carlo. The size of $H_{\rm bc}$ diminishes at larger optical depths, following a predictable $(a\tau_0)^{-1/3}$ scaling. Introducing time-dependence back into the transfer equation leads to another solution which is obtained with a discretized eigenfunction expansion and solved numerically. We demonstrate that it agrees with the steady-state result when integrated over time, though its rate of numerical convergence is quite slow and requires a sum over many modes to become accurate. The time-dependent solution is utilized to create wait-time distributions for photons escaping the sphere of optically-thick hydrogen gas, which can be sampled in lieu of direct Monte Carlo simulations that would ordinarily be computationally demanding. We compare the calculations from the time-dependent solution with Monte Carlo for a sample of line-center optical depths, noting general agreement in the resulting escape time distributions. In future work, these distributions will be fitted and used to accelerate Monte Carlo \lya transfer at large optical depths with potential applications in radiation hydrodynamic simulations of the atmospheres of exoplanets.

\acknowledgments

\appendix
\restartappendixnumbering
\section{ derivation of the transfer equation } \label{app:rteqn_derivation}

@@

>>>>

Numerical simulations can be used to understand the complex layers of these irradiated exoplanet atmospheres. However, the extreme optical depths at \lya line center introduce a computational barrier to simulating radiative transfer via Monte Carlo methods. The main impediment to performing magnetohydrodynamic simulations with fully a coupled radiative transfer solution via Monte Carlo methods is the prohibitive computational cost. The number of scatterings a photon undergoes is proportional to the line center optical depth, $\tau_0$.  Near the base of the atmosphere, the line center optical depth is ${\sim}10^9$, so most of the time is spent following photons in these cells. A method that can treat these zones via a simple solution rather than Monte Carlo has the potential to greatly accelerate the calculation. A complete, general analytic solution for the resonance scattering has never been derived but approximate solutions exist in certain limits.

@@

Starting with the full transfer equation, Eq. \ref{eq:finaleqn}, with no photon destruction and a photon emission term given by Eq. \ref{eq:jem}, the steady-state transfer equation is
\be
\nabla^2 J + \left( \frac{k}{\Delta} \right)^2 \frac{\partial^2 J}{\partial \sigma^2} & = & 
- \frac{ \sqrt{6} kL}{4\pi \Delta^2} \delta^3(\vec{x} - \vec{x}_s) \delta (\sigma - \sigma_{\rm s}).
\label{eq:rt_no_destr}
\ee
where $\sigma$ comes from a change of variables in photon frequency (Eq. \ref{eq:change_of_variables}) and $\sigma_s$ is the photon frequency of the source. The boundary condition is that there is no incoming intensity at the surface, which can be written
\citep{1986rpa..book.....R}
\be
J & = & \sqrt{3} H
\label{eq:bc}
\ee
at $r=R$.

@@

\subsection{Scaling with Line Center Optical Depth $\tau_0$}
To examine the relative size of the correction term $H_{\rm bc}$ as compared to $H_0$, we consider the case where the $\tau_0$ is large and thus $J_{\rm bc} \gg H_{\rm bc}$. We obtain the proportionality
\be \label{eq:hbc_scaling}
\frac{H_{\rm bc}(R, \sigma)}{H_0(R, \sigma)} \propto \frac{1}{(a\tau_0)^{1/3}}.
\ee
Thus, the fractional importance of the correct frequency-dependent boundary condition term $H_{\rm bc}$ scales as $(a\tau_0)^{-1/3}$. At very large $\tau_0$, it is expected that the correction term will be small, but will become increasingly important as $\tau_0$ decreases.

The location of the spectral peak of the distribution is
\be \label{eq:tau_peak_scaling}
x_{\rm peak} {\sim} \left(a\tau_0\right)^{1/3}.
\ee
The solutions of the transfer equation will perform best when the peaks of the spectral energy distribution lie well outside of the Doppler core, e.g., for large $\tau_0$. For \lya and H atoms with $T=10^4$ K, the damping parameter is $a = \Gamma / (4\pi\Delta) = 4.72\times 10^{-4}$. Thus, the value of $x$ at which the Doppler and Lorentzian components of the line wing are equal is $x_{\rm cw}=3.3$. Setting $x_{\rm cw} = x_{\rm peak}$ and solving for $\tau_0$ gives the value of the line-center optical depth at which the core of the line moves into the peak of the spectrum, which is $\tau_{\rm cp}=7.3\times10^4$. Solutions derived assuming a Lorentzian line profile $\phi(x) \approx a/(\pi x^2)$ are expected to fail for lower $\tau_0$ where the peak of the spectrum lies near the core and improve for larger $\tau_0$ where the spectral peak is well outside of the line core.

\subsection{Comparison to Monte Carlo}
The Monte Carlo method is used to solve the transfer equation numerically. This method is valid at arbitrarily large $\tau_0$, being restricted only by its computational demand which grows proportionally to the number of photons used and $\tau_0$. For each simulation, a total of ${\sim}10^6$ photon packets are initialized at a monochromatic source frequency $x_s$ and are allowed to propagate through the sphere until escaping, at which point their positions, outgoing angles, and escape frequencies are stored. A constant temperature of $T=10,000\ \rm K$ is set for the gas. Frequency redistribution is calculated at each scattering, including the effects of recoil, to obtain the spectrum at the surface of the spherical simulation domain. In the comparisons shown in this section, the raw photon data is binned in frequency to obtain spectra. 64 bins are used to ensure good statistics and minimal noise in each bin. Further details of the Monte Carlo implementation are discussed in \cite{2017ApJ...851..150H}.

\ifx
\begin{figure}
    \centering
    \includegraphics{taubc.pdf}
    \caption{Boundary condition term $H_{\rm bc}$ scaled by $(a\tau_0)^{1/3}$ shown for increasing $\tau_0$. The Monte Carlo data minus the $H_0$ solution should be close to $H_{\rm bc}$ according to Eq. \ref{eq:totalflux}. As $\tau_0$ becomes larger, the agreement between these two improves, indicating that at large optical depths the relative size of the corrective boundary condition term scales as $(a\tau_0)^{-1/3}$}.
    \label{fig:taubc}
\end{figure}
\fi

@@

Figure \ref{fig:xinit} shows quantitative examination of how the boundary term $H_{\rm bc}$ scales with both $\tau_0$ and $x_s$, motivated by the previous figures' demonstration that $H_0$ appears to become more accurate as $\tau_0$ increases. In this figure, the photon initialization frequency $x_s$ is shifted by integer multiples of $\tau_0$ in $\sigma$ frequency units (Eq. \ref{eq:change_of_variables}) such that the source, now written $\sigma_s$, falls near the peak of the spectrum for each $\tau_0$. For clarity, only the $x > 0$ side of the spectrum is shown. From Eq. \ref{eq:hbc_scaling}, it is expected that the size of $H_{\rm bc}$ should become smaller with larger optical depths following a $(a\tau_0)^{-1/3}$ scaling as compared to $H_0$. This factor has been scaled out of the figure such that solutions for different optical depths and the same source frequency $\sigma_s$ should show close agreement in scale on the figure's vertical axis if the relation holds. Indeed, the scaled solutions converge as the optical depth becomes larger, indicating general agreement with the $(a\tau_0)^{-1/3}$ scaling. The remaining discrepancy present in the vertical axis for fixed $\sigma_s$ is due to the width of the Doppler core with respect to the peak of the spectrum; at lower $\tau_0$, this causes the line profile approximation in the wing (Eq. \ref{eq:line_profile_wing}) to break down. From this, we conclude that the errors introduced by the incorrect separation of variables in \cite{1973MNRAS.162...43H}, \cite{1990ApJ...350..216N}, \cite{2006ApJ...649...14D} and others are proportional to $(a\tau_0)^{-1/3}$.

@@

Let us define the damping frequency to be $\gamma \equiv i\omega$, which is real-valued and positive for damped solutions to the time-dependent equation. There are certain values of $\gamma$ for which a resonant response is produced which we will call $\gamma_{nm}$. Near these $\gamma_{nm}$ poles, the eigenfunctions will have the form 
\be \label{eq:jnsigmaomega_approx}
J(n,\sigma,-i\gamma) & \simeq \frac{ J_{nm}(\sigma) }{\gamma_{nm} - \gamma} + {\rm terms\ which\ vary\ slowly\ in\ \gamma}.
\ee
From Eq. \ref{eq:jrsigmat_expansion}, the contribution of a single pole to the total solution is
\be
J_{nm}(r,\sigma,t) & = &  -j_0(\kappa_n r)  \int \frac{d\gamma}{2\pi i} e^{-\gamma t}J(n,\sigma,-i\gamma)
\\ & \simeq & j_0(\kappa_n r)  \int \frac{d\gamma}{2\pi i} e^{-\gamma t} \frac{ J_{nm}(\sigma) }{\gamma - \gamma_{nm}}.
\ee
We close the integral along the imaginary $\gamma$-axis  with a semi-circle at infinity and apply the residue theorem. The contribution from the semi-circle is zero, so we obtain
\be
J_{nm}(r,\sigma,t) & \simeq & j_0(\kappa_n r) J_{nm}(\sigma) e^{-\gamma_{nm}t}.
\ee
Summing over all spatial modes $n$ and over all eigenmodes $m$ for a given $n$, we obtain
\be \label{eq:Jrsigmat}
J(r,\sigma,t) & = & \sum_{n=1}^\infty j_0(\kappa_n r)  \sum_{m=1}^{\infty} J_{nm}(\sigma) e^{-\gamma_{nm}t}.
\ee
This ansatz captures the contributions from $n \times m$ simple poles, which seems to agree with Monte Carlo results for $x_s=0$ based on numerical calculations of this result. However, the analysis presented in Appendix \ref{app:wkb} suggests the ansatz is incomplete in the case where $x_s \neq 0$ and the spectrum is asymmetric.

@@

The form of a single eigenmode $J_{nm}(\sigma)$ is oscillatory out to some turning point, $\sigma_{tp}$, at which point the function becomes evanescent. The location of the turning point can be found by ignoring the delta-function discontinuity at the source frequency $\sigma_s$ in Eq. \ref{eq:diffusion_plugged_in} and re-examining the resulting homogeneous differential equation. We obtain
\be
\frac{d^2J}{d\sigma^2} & = & \left[ \left( \frac{\kappa_n \Delta }{k} \right)^2 - \frac{3\phi \gamma\Delta^2}{ck}\right] J,
\ee
where the line profile is approximated as in Eq. \ref{eq:line_profile_wing}. When the coefficient on the right hand side is positive, exponential growth or decay solutions are found (evanescence). This occurs on the line wings. When the coefficient on the right hand side is negative, oscillatory solutions are found (propagation), which occurs near the line core. The boundary between propagation and evanescence occurs at the turning point, given by
\be
\sigma_{tp} & = & \sqrt{\frac{2a}{\pi}}\left( \frac{k \gamma}{ \kappa_n^2 c \Delta} \right)^{3/2}
\ee
Thus, to ensure accuracy in the final sum of Eq. \ref{eq:Jrsigmat}, the bounds of $\sigma$ must be set sufficiently far outside of $\sigma_{tp}$ such that the function is small at the edges. The scale of an $e$-folding in $J_{nm}(\sigma)$ is $k/(\kappa_n \Delta) = \tau_0 / (\sqrt{\pi} n)$, so a grid of $\sigma$ is chosen that spans a large enough number of $e$-foldings that no oscillatory behavior is present at the boundaries of the calculation.

@@

In Fig. \ref{fig:tau_scaling}, it is shown that the time constant of exponential decay in fitted Monte Carlo escape time distributions converges with $\gamma_{00}^{-1}$ at sufficiently high $\tau_0$, following a $t\propto(a\tau_0)^{1/3}$ scaling. At lower $\tau_0$, the effects of line core scattering are most important, leading to a larger discrepancy in the characteristic escape timescale. Here, the Monte Carlo accurately characterizes the photons which scatter in the core many times before escaping, while the semi-analytic solution does not capture this behavior as it uses only the Lorentzian piece of the line profile and does not use enough spatial modes to accurately model the frequency regime near line center. However, as $\tau_0$ grows, the effect of core scattering becomes smaller and the approximations hold, agreeing well with the expected $(a\tau_0)^{1/3}$ scaling for the rate constant.

\begin{figure}
    \centering
    \includegraphics[width=0.75\textwidth]{tau_scaling.pdf}
    \caption{Characteristic timescales as a function of $\tau_0$. $\gamma_{00}^{-1}$ and a fit of the exponential falloff for the Monte Carlo escape time distribution are shown to converge to a $t \propto (a\tau_0)^{1/3}$ scaling at large $\tau_0$. The Monte Carlo points on this figure are obtained by fitting only the exponential piece of the escape time distribution to obtain the rate constant.}
    \label{fig:tau_scaling}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics{waittime.pdf}
    \caption{Wait time distributions of escaping photons for $\tau_0=10^6$ and $10^7$, which is fit using Eq. \ref{eq:fitting_function}. A sum over 20 spatial eigenmodes and 500 frequency eigenmodes is labeled ``Eigenfunctions''. All calculations were performed with a monochromatic source of photons at line center ($x_s = 0.0$).}
    \label{fig:escape_time}
\end{figure}

@@

Models of heating and escape of hydrogen gas in exoplanet atmospheres can be constructed with an accurate treatment of resonant scattering of \lya in spherical geometry. The Monte Carlo method can be used for this problem, but requires a solution of radiative transfer in a sphere and is limited by its high computational demand for large $\tau_0$ where there are many photon scatterings before escape. The primary goal of this work is to present a solution for mean intensity ($J$) and flux ($H = F/4\pi$) of photons near line-center frequency $\nu_0$ in a uniform sphere. We have generalized a solution derived by \cite{2006ApJ...649...14D} (called $H_0$ here) to allow a monochromatic source of photons with frequencies away from line center, and we have made a correction to their incorrect $J=0$ surface boundary condition. This correction allows the proper condition $J=\sqrt{3}H$ to be satisfied at the surface of the sphere, and comes from the introduction of the terms $J_{\rm bc}$ and $H_{\rm bc}$ which are solved using a continuous Fourier expansion in frequency. Enforcement of the correct boundary term leads naturally to expansion of the spatial dependence in terms of spherical Bessel functions. In practice, the continuous expansions are discretized and the Fourier coefficients solved for via a matrix equation. The resulting flux correction, $H_{\rm bc}$, scales as $(a\tau_0)^{-1/3}$. Thus, for very large $\tau_0$, only a small correction to $H_0$ is needed, while larger errors are present in calculations performed at lower $\tau_0$. Since the Laplacian form of the differential equation is only attainable by approximating the line profile as $\phi \approx a/(\pi x^2)$, these solutions are invalid in areas where the Doppler core of the \lya line has a significant effect on the scattering of photons. The peak of the spectral energy distribution of escaping photons is $x_0 = (a\tau_0)^{1/3}$; as such, calculations performed at low $\tau_0$ are likely to be inaccurate due to the close proximity of the spectral peak and the Doppler core of the line. By comparison with Monte Carlo simulations, we have shown that the enforcement of the correct frequency-dependent boundary conditions improves the accuracy of these analytic solutions %and correctly tracks changes in the spectrum of escaping photons as the monochromatic source frequency $x_s$ is altered, 
provided a sufficiently large $\tau_0$. Specifically, this solution shows improvement over previous solutions that utilized a $J=0$ surface boundary condition, such as \cite{1973MNRAS.162...43H}, \cite{1990ApJ...350..216N}, \cite{2006ApJ...649...14D} and others. TODO: Give details. The surface boundary condition correction $H_{\rm bc}$ is positive in the line wing and negative at the peak of the spectrum, which suggests the previously cited solutions are too large at the spectral peaks and too small further out in the wing, where the error scales approximately as $(a\tau_0)^{-1/3}$. This appears to be the consequence of not enforcing the correct frequency-dependent boundary conditions at the surface of the medium.

@@

We have examined previous solutions to \lya transfer in the limit of large optical depth, noting that the separation of variables and treatment of the boundary condition in \cite{1973MNRAS.162...43H}, \cite{1990ApJ...350..216N}, \cite{2006ApJ...649...14D} and others does not produce a valid solution to the transfer equation. Here, we have derived the solution in spherical geometry with an appropriate treatment of the surface boundary condition as a frequency-dependent quantity, examining the spectral and time dependence of the resulting equations. The key result is that the errors in the previously-cited works have been quantified via a corrective boundary term, $H_{\rm bc}$, which resolves an excess in flux at the spectral peak and a deficit in the line wing of the calculated spectrum of \lya radiation as compared with Monte Carlo. The size of $H_{\rm bc}$ diminishes at larger $\tau_0$, following a predictable $(a\tau_0)^{-1/3}$ scaling. Introducing time-dependence back into the transfer equation leads to another solution which is obtained with a discretized eigenfunction expansion and solved numerically. We demonstrate that it agrees with the steady-state result when integrated over time, though its rate of numerical convergence is quite slow and requires a sum over many modes to become accurate. The time-dependent solution is utilized to create wait-time distributions for photons escaping the sphere of optically-thick hydrogen gas, which can be sampled in lieu of direct Monte Carlo simulations that would ordinarily be computationally demanding. We compare the calculations from the time-dependent solution with Monte Carlo for a sample of $\tau_0$, noting general agreement in the resulting escape time distributions. In future work, this will be used to accelerate Monte Carlo \lya transfer at large optical depths with potential applications in radiation hydrodynamic simulations of the atmospheres of exoplanets.

\acknowledgments
\restartappendixnumbering

\appendix
\section{ derivation of the transfer equation } \label{app:rteqn_derivation}

@@

\section{WKB Approximation} \label{app:wkb}
